Programmer friendly communication framework

### 2024-03-02 2024_paper/paper/msglink_framework.tex
added formatter info
cleaned up style of document

\begin{abstract}
This document will focus on an important topic of cybersecurity: Passwords. It will have a look on the indispensable role of passwords now and in the future. Additionally it will cover the role of passwords in robotics and will provide a best practice guide for working with password safes and how to create a secure and easy to remember master passphrase.
\end{abstract}

\section{Introduction}
The present paper concentrates on the wide use of passwords or passphrases in IT-security, with a special consideration of robotics. It focuses on the security aspect of passwords, especially on weak passwords due to user (human) laziness, which in some cases can lead to major problems. \\
Further, it covers state of the art password technologies (2 factor authentication with passkeys) and possible future technologies.\\
In addition the review highlights alternatives for effective password management and it introduces a choice of selected password managing software. \\
Finally the paper includes a best practice guide to:
\begin{itemize}
\item check if user's passwords have already been breached 
\item create an easy to remember but secure master password for a password managing software
\end{itemize}

\section{Implications of weak passwords}
In today's digital world many aspects of our daily lives rely on remote services, like online banking, social media and smart home devices to name just a few of them. The majority of those services require the user to authenticate him- or herself via username and password. A weak or re-used password could become a major security issue, if an account gets breached it can have devastating affects on the account owner's life.\\   
But not only humans can be negatively affected. Many robots in today's world are controlled or maintained remotely. To regulate access, users will need a high security password. Especially in robotics, using a weak one can cause major problems, because users don't want anybody to have access to their robots.

@@

\section{State of the art}
Sensitive information or access to hardware must be protected in some way. Nowadays this is likely to happen with a username, which is normally easy to guess, and a password, which should be complex and secure. Also the data traffic should be encrypted, so nobody can read along and possibly gain access to it.\\
For example users need to key in a username and password before they can access any network device over SSH. This is also the case if a client wants to access the Raspberry Pi inside the Wombat controller remotely. Over SSH the data traffic is encrypted and protected with a password.

\subsection{The problem with passwords} 
As mentioned above, a password protects sensitive information or access to hardware like a robot. That's why it's so important to have a good, secure password. However users often are too lazy to create a strong password for every account they own. This is understandable, because as often cited study from NordPass shows, the average user has around 100 accounts that require a unique password. Remembering a hundred unique passwords or more is really beyond a users capability. In a matter of that, users often use the same password for all programs and devices. For example if a user owns 150 robots and all of them have the same password a hacker only needs to find out one and already has access to all of his or her robots. If a user is using 150 different password, a successful hacker will just be able to control one robot by one hack.\cite{b1} \cite{b2}\\
Beside laziness as the main cause there are many other reasons why users tend to create weak passphrases such as lack of risk awareness or the fear of forgetting it.\\
Every year long records get published by different providers revealing the most used passwords. Here are the top ten used passwords of last year (2022): \cite{b3}
\begin{enumerate}
\item 123456
\item 123456789
\item qwerty
\item password
\item 12345
\item 12345678
\item 111111
\item 1234567
\item 123123
\item 1234567890
\end{enumerate}
Other common passwords are locations, names, birthdays, etc.

@@

\subsection{Current state of the art}
In recent years there was a massive increase in cybercrime. To prevent data-leaks, companies started integrating 2-factor-authentication.\cite{b4}\\
2-factor-authentication adds an extra layer of security to passwords. First the user has to fill in the username and the password like always. Then the user has to pass an additional login-method to verify it's really him or her.\\
The second factor is usually divided into one of the following categories:
\begin{itemize}
\item \textbf{Something only the user knows.}  This might be a personal question the user had to configure beforehand, a personal identification number (PIN), etc.
\item \textbf{Something only the user possesses.} This could be a small hardware token or smart-phone.
\item \textbf{Something physical unique to the user.} This might be a  biometric pattern like a fingerprint, an iris scan, or a voice print. 
\end{itemize}
The most common type of the 2-factor-authentication is a smartphone. Another good - but less common - option is a hardware key. (Something only the user possesses.) A leading company in this market is yubico. They are providing a security key named YubiKey for devices with USB-A, USB-C, Lightning and NFC. Users that have a YubiKey should be aware that they have to carry it with them at all times and be careful not to lose it.\cite{b11}\\
Another widely used method is the SMS Text-Message. (Something only the user knows.) After entering the password and username the user receives a unique one time passcode (OTP) via text message. The user needs to key in the passcode during a limited time (for example one minute).\\
After both the first factor and the second factor are correct the user gains access.

\subsection{Future state of the art}
In the future passwords will likely be entirely replaced by passkeys because they are easier to use, securer, faster and still work on almost every device.\\
A passkey is similar to the 2-factor-authentication but without the first factor. The second and only factor is the user's device's security method such as a PIN or a biometric sensor (fingerprint, face scan). \\
The passkey is hardware specific, which means the user always needs to carry a second device (that the passkey is installed on). If the user wants to use a passkey on the laptop he or she needs to verify the identity with an additional smartphone. This eliminates phishing, credential stuffing and other remote attacks. \\
In conclusion when a user is asked to sign-in to an app or website, he or she approves the sign-in with the same biometric or PIN that he or she uses to unlock the device (phone, computer or security key). The app or website can use this mechanism instead of the traditional (and insecure) username and password. \cite{b5}

\subsection{Solutions to password managing}
If users don't want to or are not able to use passkeys there are different options how they could protect their sensitive information. Users can write their passwords with their account information into a booklet. Thus, the passwords can be unique and strong. It's a really simple idea but also has a lot of drawbacks. 
For example:
\begin{itemize}
\item Users need to always carry the booklet with them. Without it they are not able to access their accounts.
\item Users can lose the booklet. Depending on the security restrictions they can't gain access to some of their accounts anymore.
\item It's not a really clear way of organizing passwords. Once something is written down it cannot change the position in the booklet. So users would have to search for each password for indefinite time before they find it.
\end{itemize}
The other and better option is a password safe.

\subsection{Password Safe}
\subsubsection{Explanation}
A password safe, also known as password manager, is a piece of sofware that allows saving a list of usernames and passwords for all accounts. It has one master password or passphrase that unlocks the safe. Now users only have to recall one single master passphrase instead of a huge amount of passwords.
\subsubsection{Examples}
\begin{itemize}
\item \textbf{KeepassXC}\\ 
\begin{center}
\includegraphics[scale=0.6]{./images/KeePass.png}
\captionof{figure}[KeePassXC - The figure is taken from the homepage of the KeePassXC Project (https://keepassxc.org/)]{KeePassXC Logo}
\end{center}

@@

\item \textbf{1Password}\\
\begin{center}
\includegraphics[scale=0.06]{./images/1Password-Logo.wine.png}
\captionof{figure}[1Password - The figure is taken from the homepage of the 1Password Project (https://1password.com/)]{1Password Logo}
\end{center}
This password safe is known for its easy to use interface and its high security encryption.\\ 
1Password uses an uncommon encryption known as dual-key encryption. If the server gets breached it's impossible for the hacker to decrypt the user's sensitive information because of the two required keys. The first key is the user's master-key. The second key is a secret key, which is a 128-bit, machine-generated code. The secret key is generated on every device the user logs into. It will only be saved on the user's devices and will never be saved together with the second PINs.\\
Depending on the account type the data is stored differently, but there is only one account type saving the users' data locally. In every other option the user's data gets saved in a cloud-based vault.\\
The interface is known as user-friendly and generates strong passwords for accounts that are newly created.
\cite{b7}

\end{itemize}
\subsubsection{Benefits and drawbacks of password safe software}
\textbf{Benefits}:
\begin{itemize}
\item \textbf{Security:} Password safes use advanced encryption algorithms to protect the user's passwords. This makes it extremely difficult for hackers to steal the passwords and gain access to the user's online accounts.
\item \textbf{Convenience:} Password safes store all of the user's passwords in one place, which makes it easy for them to access their accounts without remembering multiple passwords.
\item \textbf{Auto-fill feature:} Many password safes have an auto-fill feature that automatically enters the username and the password of the user, saving time and reducing the risk of typing errors.
\item \textbf{Multi-device access:} Many password safes allow the user to access their passwords from multiple devices, including smartphones, tablets, and computers.
\item \textbf{Password strength:} Password safes often include a password generator feature, which creates strong passwords. This helps to ensure that the passwords are not easily guessable.
\end{itemize}
\textbf{Drawbacks}:
\begin{itemize}
\item \textbf{Single point of failure:} If a user's password safe is hacked or compromised, all of their passwords are at risk. It is important to choose a password safe that has a strong encryption algorithm and to use a strong master password.
\item \textbf{Dependency:} Using a password safe can make a user dependent on it for remembering passwords, which may make it difficult to remember passwords if the user is not able to access their password safe for some reason.
\item \textbf{Learning curve:} Some password safes have a steep learning curve, and it can take some time to get used to the application.
\item \textbf{Cost:} Some password safes charge a fee for their services, which can be a disadvantage if the user is looking for a free solution.
\item \textbf{Compatibility:} Some password safes may not be compatible with all websites or devices, which can limit their usefulness.
\end{itemize}
Sources used to find the benefits and drawbacks: \cite{b8}
\subsubsection{Scientific Studies regarding password managers}
A thoughtful scientific study by Peter Mayer et al. shows, that even for highly educated people, which definitely know about the risk of bad or reused passwords, the adoption of password manager software depends largely on how easy it is to use. Mayer and his colleagues show, that browser based password managers are more widely used than standalone password saves, mainly because they are easier to use.\cite{b13}

\section{Concept and implementation for best practice}
It can be very overwhelming to start with a secure option to manage passwords. So here is a best practice guide everyone can follow.
\begin{enumerate}
\item First you should check if your password has already been breached. There is a website that makes it easy to check it out. It's called \textbf{"haveibeenpwned.com"}.\\
Haveibeenpwned is open source and secure. If someone searches for a breach it doesn't store the data that was typed in (an email address or phone number). It only ever retrieves the data from storage then returns the data. \vspace{0.5cm}
\begin{center}
\includegraphics[scale=0.29]{./images/haveibeenpwned.png}
\captionof{figure}[Have I been pwned - The figure is taken from the homepage of the HaveIBeenPwned? Project (https://haveibeenpwned.com/)]{Have I been pwned? Homepage}
\end{center}
If you have already been pawned, change the present password. If not you can rest assured. 
\item Pick a password safe that fits best for your needs. You can choose one of the ones mentioned above or do your own research.
\item Create a good master password or even better a master passphrase. Here is a easy way to to design it: 
\begin{itemize}
\item First think of something which is easy to remember. For example "I love apples and bananas" or "Ich komme aus der Stadt Wien" if you want to use your native language.
\item The second step is to transform the sentence in dialect or slang form. For example "I luv appls and bananas" or "I komm aus da Stadt Wien".
\item Third, you can change letters into numbers and remove spaces. A "s" gets changes into a "5", an "e" gets changed into a "3" and an "o" gets changed into a "0". For example "Iluvappl5andbanana5" or "Ik0mmau5da5tadtWi3n".
\item The fourth step is to add symbols to your password. Change characters that look like a symbol into one such as "a" to "@" or "s" to "?". For example "Iluv@ppl5@ndb@n@n@5" or "Ikomm@u5d@5tadtWi3n".
\item The fifth step is to make all words start with a capital letter. This is only useful if the language you choose does not have a grammar rule for that (It's needed in English but not in German for example). 
\end{itemize}
Finally you would end up with a passphrase like "ILuv@ppl5@ndB@n@n@5" or "Ikomm@u5d@5tadtWi3n". This is a really secure passphrase because it compromises numbers, upper case and lower case letters, symbols and more than 8 characters. This would make a strong master password that is nearly impossible to crack.
\end{enumerate}

\section{Conclusion}
Passwords remain a widely used and convenient method of authentication, but they have significant weaknesses that make them vulnerable when attacked. While there are emerging alternatives to passwords, such as biometric authentication and passkeys, there are also challenges associated with their adoption, such as cost, compatibility, and user experience.\\
As the cybersecurity landscape continues to evolve, it is important to explore new, innovative and equally trustworthy approaches to authentication and access control, that are both secure and user-friendly.

%Abbildungsverzeichnis erstellen
\listoffigures

>>>>

\begin{abstract}
    This document will focus on an important topic in software development: Communication. It will provide an insight into why what communication is needed for and how communication between computers is typically realized. Additionally it will cover problems and annoyances with currently available communication frameworks for certain applications as well as introduce concepts and a possible solution for the before-mentioned. 
\end{abstract}

\section{Introduction}

The present paper concentrates on the wide use of passwords or passphrases in IT-security, with a special consideration of robotics. It focuses on the security aspect of passwords, especially on weak passwords due to user (human) laziness, which in some cases can lead to major problems. \\
Further, it covers state of the art password technologies (2 factor authentication with passkeys) and possible future technologies.\\
In addition the review highlights alternatives for effective password management and it introduces a choice of selected password managing software. \\
Finally the paper includes a best practice guide to:
\begin{itemize}
    \item check if user's passwords have already been breached
    \item create an easy to remember but secure master password for a password managing software
\end{itemize}

\section{Implications of weak passwords}

In today's digital world many aspects of our daily lives rely on remote services, like online banking, social media and smart home devices to name just a few of them. The majority of those services require the user to authenticate him- or herself via username and password. A weak or re-used password could become a major security issue, if an account gets breached it can have devastating affects on the account owner's life.\\
But not only humans can be negatively affected. Many robots in today's world are controlled or maintained remotely. To regulate access, users will need a high security password. Especially in robotics, using a weak one can cause major problems, because users don't want anybody to have access to their robots.

@@

\section{State of the art}

Sensitive information or access to hardware must be protected in some way. Nowadays this is likely to happen with a username, which is normally easy to guess, and a password, which should be complex and secure. Also the data traffic should be encrypted, so nobody can read along and possibly gain access to it.\\
For example users need to key in a username and password before they can access any network device over SSH. This is also the case if a client wants to access the Raspberry Pi inside the Wombat controller remotely. Over SSH the data traffic is encrypted and protected with a password.

\subsection{The problem with passwords}

As mentioned above, a password protects sensitive information or access to hardware like a robot. That's why it's so important to have a good, secure password. However users often are too lazy to create a strong password for every account they own. This is understandable, because as often cited study from NordPass shows, the average user has around 100 accounts that require a unique password. Remembering a hundred unique passwords or more is really beyond a users capability. In a matter of that, users often use the same password for all programs and devices. For example if a user owns 150 robots and all of them have the same password a hacker only needs to find out one and already has access to all of his or her robots. If a user is using 150 different password, a successful hacker will just be able to control one robot by one hack.\cite{b1} \cite{b2}\\
Beside laziness as the main cause there are many other reasons why users tend to create weak passphrases such as lack of risk awareness or the fear of forgetting it.\\
Every year long records get published by different providers revealing the most used passwords. Here are the top ten used passwords of last year (2022): \cite{b3}
\begin{enumerate}
    \item 123456
    \item 123456789
    \item qwerty
    \item password
    \item 12345
    \item 12345678
    \item 111111
    \item 1234567
    \item 123123
    \item 1234567890
\end{enumerate}
Other common passwords are locations, names, birthdays, etc.

@@

\subsection{Current state of the art}

In recent years there was a massive increase in cybercrime. To prevent data-leaks, companies started integrating 2-factor-authentication.\cite{b4}\\
2-factor-authentication adds an extra layer of security to passwords. First the user has to fill in the username and the password like always. Then the user has to pass an additional login-method to verify it's really him or her.\\
The second factor is usually divided into one of the following categories:
\begin{itemize}
    \item \textbf{Something only the user knows.}  This might be a personal question the user had to configure beforehand, a personal identification number (PIN), etc.
    \item \textbf{Something only the user possesses.} This could be a small hardware token or smart-phone.
    \item \textbf{Something physical unique to the user.} This might be a  biometric pattern like a fingerprint, an iris scan, or a voice print.
\end{itemize}
The most common type of the 2-factor-authentication is a smartphone. Another good - but less common - option is a hardware key. (Something only the user possesses.) A leading company in this market is yubico. They are providing a security key named YubiKey for devices with USB-A, USB-C, Lightning and NFC. Users that have a YubiKey should be aware that they have to carry it with them at all times and be careful not to lose it.\cite{b11}\\
Another widely used method is the SMS Text-Message. (Something only the user knows.) After entering the password and username the user receives a unique one time passcode (OTP) via text message. The user needs to key in the passcode during a limited time (for example one minute).\\
After both the first factor and the second factor are correct the user gains access.

\subsection{Future state of the art}

In the future passwords will likely be entirely replaced by passkeys because they are easier to use, securer, faster and still work on almost every device.\\
A passkey is similar to the 2-factor-authentication but without the first factor. The second and only factor is the user's device's security method such as a PIN or a biometric sensor (fingerprint, face scan). \\
The passkey is hardware specific, which means the user always needs to carry a second device (that the passkey is installed on). If the user wants to use a passkey on the laptop he or she needs to verify the identity with an additional smartphone. This eliminates phishing, credential stuffing and other remote attacks. \\
In conclusion when a user is asked to sign-in to an app or website, he or she approves the sign-in with the same biometric or PIN that he or she uses to unlock the device (phone, computer or security key). The app or website can use this mechanism instead of the traditional (and insecure) username and password. \cite{b5}

\subsection{Solutions to password managing}

If users don't want to or are not able to use passkeys there are different options how they could protect their sensitive information. Users can write their passwords with their account information into a booklet. Thus, the passwords can be unique and strong. It's a really simple idea but also has a lot of drawbacks.
For example:
\begin{itemize}
    \item Users need to always carry the booklet with them. Without it they are not able to access their accounts.
    \item Users can lose the booklet. Depending on the security restrictions they can't gain access to some of their accounts anymore.
    \item It's not a really clear way of organizing passwords. Once something is written down it cannot change the position in the booklet. So users would have to search for each password for indefinite time before they find it.
\end{itemize}
The other and better option is a password safe.

\subsection{Password Safe}

\subsubsection{Explanation}

A password safe, also known as password manager, is a piece of sofware that allows saving a list of usernames and passwords for all accounts. It has one master password or passphrase that unlocks the safe. Now users only have to recall one single master passphrase instead of a huge amount of passwords.

\subsubsection{Examples}

\begin{itemize}
    \item \textbf{KeepassXC} \\
        \begin{center}
            \includegraphics[scale=0.6]{./images/KeePass.png}
            \captionof{figure}[KeePassXC - The figure is taken from the homepage of the KeePassXC Project (https://keepassxc.org/)]{KeePassXC Logo}
        \end{center}

@@

    \item \textbf{1Password} \\
        \begin{center}
            \includegraphics[scale=0.06]{./images/1Password-Logo.wine.png}
            \captionof{figure}[1Password - The figure is taken from the homepage of the 1Password Project (https://1password.com/)]{1Password Logo}
        \end{center}
        This password safe is known for its easy to use interface and its high security encryption. \\
        1Password uses an uncommon encryption known as dual-key encryption. If the server gets breached it's impossible for the hacker to decrypt the user's sensitive information because of the two required keys. The first key is the user's master-key. The second key is a secret key, which is a 128-bit, machine-generated code. The secret key is generated on every device the user logs into. It will only be saved on the user's devices and will never be saved together with the second PINs.\\
        Depending on the account type the data is stored differently, but there is only one account type saving the users' data locally. In every other option the user's data gets saved in a cloud-based vault.\\
        The interface is known as user-friendly and generates strong passwords for accounts that are newly created.
        \cite{b7}
\end{itemize}

\subsubsection{Benefits and drawbacks of password safe software}

\textbf{Benefits}:
\begin{itemize}
    \item \textbf{Security:} Password safes use advanced encryption algorithms to protect the user's passwords. This makes it extremely difficult for hackers to steal the passwords and gain access to the user's online accounts. 
    \item \textbf{Convenience:} Password safes store all of the user's passwords in one place, which makes it easy for them to access their accounts without remembering multiple passwords.
    \item \textbf{Auto-fill feature:} Many password safes have an auto-fill feature that automatically enters the username and the password of the user, saving time and reducing the risk of typing errors.
    \item \textbf{Multi-device access:} Many password safes allow the user to access their passwords from multiple devices, including smartphones, tablets, and computers.
    \item \textbf{Password strength:} Password safes often include a password generator feature, which creates strong passwords. This helps to ensure that the passwords are not easily guessable.
\end{itemize}

\textbf{Drawbacks}:
\begin{itemize}
    \item \textbf{Single point of failure:} If a user's password safe is hacked or compromised, all of their passwords are at risk. It is important to choose a password safe that has a strong encryption algorithm and to use a strong master password.
    \item \textbf{Dependency:} Using a password safe can make a user dependent on it for remembering passwords, which may make it difficult to remember passwords if the user is not able to access their password safe for some reason.
    \item \textbf{Learning curve:} Some password safes have a steep learning curve, and it can take some time to get used to the application.
    \item \textbf{Cost:} Some password safes charge a fee for their services, which can be a disadvantage if the user is looking for a free solution.
    \item \textbf{Compatibility:} Some password safes may not be compatible with all websites or devices, which can limit their usefulness.
\end{itemize}

Sources used to find the benefits and drawbacks: \cite{b8}

\subsubsection{Scientific Studies regarding password managers}

A thoughtful scientific study by Peter Mayer et al. shows, that even for highly educated people, which definitely know about the risk of bad or reused passwords, the adoption of password manager software depends largely on how easy it is to use. Mayer and his colleagues show, that browser based password managers are more widely used than standalone password saves, mainly because they are easier to use.\cite{b13}

\section{Concept and implementation for best practice}

It can be very overwhelming to start with a secure option to manage passwords. So here is a best practice guide everyone can follow.

\begin{enumerate}
    \item 
        First you should check if your password has already been breached. There is a website that makes it easy to check it out. It's called \textbf{"haveibeenpwned.com"}. Haveibeenpwned is open source and secure. If someone searches for a breach it doesn't store the data that was typed in (an email address or phone number). It only ever retrieves the data from storage then returns the data. 
        \begin{center}
            \includegraphics[scale=0.29]{./images/haveibeenpwned.png}
            \captionof{figure}[Have I been pwned - The figure is taken from the homepage of the HaveIBeenPwned? Project (https://haveibeenpwned.com/)]{"Have I been pwned?" Homepage}
        \end{center}
        If you have already been pawned, change the present password. If not you can rest assured.
    \item 
        Pick a password safe that fits best for your needs. You can choose one of the ones mentioned above or do your own research.
    \item 
        Create a good master password or even better a master passphrase. Here is a easy way to to design it:
        \begin{itemize}
            \item First think of something which is easy to remember. For example "I love apples and bananas" or "Ich komme aus der Stadt Wien" if you want to use your native language.
            \item The second step is to transform the sentence in dialect or slang form. For example "I luv appls and bananas" or "I komm aus da Stadt Wien".
            \item Third, you can change letters into numbers and remove spaces. A "s" gets changes into a "5", an "e" gets changed into a "3" and an "o" gets changed into a "0". For example "Iluvappl5andbanana5" or "Ik0mmau5da5tadtWi3n".
            \item The fourth step is to add symbols to your password. Change characters that look like a symbol into one such as "a" to "@" or "s" to "?". For example "Iluv@ppl5@ndb@n@n@5" or "Ikomm@u5d@5tadtWi3n".
            \item The fifth step is to make all words start with a capital letter. This is only useful if the language you choose does not have a grammar rule for that (It's needed in English but not in German for example).
        \end{itemize}
        Finally you would end up with a passphrase like "ILuv@ppl5@ndB@n@n@5" or "Ikomm@u5d@5tadtWi3n". This is a really secure passphrase because it compromises numbers, upper case and lower case letters, symbols and more than 8 characters. This would make a strong master password that is nearly impossible to crack.
\end{enumerate}

\section{Conclusion}

Passwords remain a widely used and convenient method of authentication, but they have significant weaknesses that make them vulnerable when attacked. While there are emerging alternatives to passwords, such as biometric authentication and passkeys, there are also challenges associated with their adoption, such as cost, compatibility, and user experience.\\
As the cybersecurity landscape continues to evolve, it is important to explore new, innovative and equally trustworthy approaches to authentication and access control, that are both secure and user-friendly.

% Abbildungsverzeichnis
\listoffigures
